From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-02 Pi+/paper/reversible1.tex
note about changing thm
Minor: formatting denotation

{\scalebox{\scalef}{$%
\begin{array}{lrcl}
\textit{Value types} & A,B,C,D &::=& \zerot \alt \onet \alt A+B \alt A\times B \\
\textit{Values}      & v,w,x,y &::=& \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
\textit{Program types} &&& A \iso B \\
\textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
\end{array}$}}
\begin{figure}[t]
{\scalebox{\scalef}{$%
%%\noindent\begin{minipage}{.7\linewidth}
\begin{array}{rrcll}
\idc :& A & \iso & A &: \idc \\
\\
\identlp :&  \zerot + A & \iso & A &: \identrp \\
\swapp :&  A + B & \iso & B + A &: \swapp \\
\assoclp :&  A + (B + C) & \iso & (A + B) + C &: \assocrp \\ [1.5ex]
\identlt :&  \onet \times A & \iso & A &: \identrt \\
\swapt :&  A \times B & \iso & B \times A &: \swapt \\
\assoclt :&  A \times (B \times C) & \iso & (A \times B) \times C &: \assocrt \\ [1.5ex]
\absorbr :&~ \zerot \times A & \iso & \zerot ~ &: \factorzl \\
\dist :&~ (A + B) \times C & \iso & (A \times C) + (B \times C)~ &: \factor
  \end{array}$}}

@@

The denotation of types is straightforward:
\[\begin{array}{rcl}
\denot{\zerot} &=& \emptyset \\
\denot{\onet} &=& \{ \Acon{tt} \} \\
\denot{A + B} &=& \denot{A} \uplus \denot{B} \\
\denot{A \times B} &=& \denot{A} \times \denot{B}
\end{array}\]
where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
sets. The denotation of a combinator $c : A \isot B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:
\[\begin{array}{rcl}
\denot{\identlp} (\inlv{v}) &=& v \\
\denot{\identrp} v &=& \inlv{v} \\
\denot{\swapp} (\inlv{v}) &=& \inrv{v} \\
\denot{\swapp} (\inrv{v}) &=& \inlv{v} \\
\denot{\assoclp} (\inlv{v})  &=& \inlv{(\inlv{v})} \\
\denot{\assoclp} (\inrv{(\inlv{v})}) &=& \inlv{(\inrv{v})} \\
\denot{\assoclp} (\inrv{(\inrv{v})}) &=& \inrv{v} \\
\denot{\assocrp} (\inlv{(\inlv{v})}) &=& \inlv{v} \\
\denot{\assocrp} (\inlv{(\inrv{v})}) &=& \inrv{(\inlv{v})} \\
\denot{\assocrp} (\inrv{v})  &=& \inrv{(\inrv{v})} \\
\denot{\identlp} (\Acon{tt} , v)  &=& v \\
\denot{\identrp} v &=& (\Acon{tt} , v) \\
\denot{\swapt}  (v_1 , v_2) &=& (v_2 , v_1) \\
\denot{\assoclt} (v_1 , (v_2 , v_3)) &=& ((v_1 , v_2) , v_3) \\
\denot{\assocrt} ((v_1 , v_2) , v_3) &=& (v_1 , (v_2 , v_3)) \\
\denot{\dist} (\inlv{v_1} , v_3)  &=& \inlv{(v_1 , v_3)} \\
\denot{\dist} (\inrv{v_2 , v_3})  &=& \inrv{(v_2 , v_3)} \\
\denot{\factor}  (\inlv{(v_1 , v_3)}) &=& (\inlv{v_1} , v_3) \\
\denot{\factor}  (\inrv{(v_2 , v_3)}) &=& (\inrv{v_2} , v_3) \\
\denot{\idc} v &=& v \\
\denot{(c_1 \fatsemi c_2)} v &=& (\denot{c_2} \circ \denot{c_1}) v \\
\denot{(c_1 \oplus c_2)} (\inlv{v}) &=& \inlv{(\denot{c_1}~v)} \\
\denot{(c_1 \oplus c_2)} (\inrv{v}) &=& \inrv{(\denot{c_2}~v)} \\
\denot{(c_1 \otimes c_2)} (v_1 , v_2) &=& (\denot{c_1} v_1 , \denot{c_2} v_2)
  \end{array}\]

@@

>>>>

{\scalebox{\scalef}{$%
    \begin{array}{lrcl}
      \textit{Value types}   & A,B,C,D & ::= & \zerot \alt \onet \alt A+B \alt A\times B        \\
      \textit{Values}        & v,w,x,y & ::= & \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
      \textit{Program types} &         &     & A \iso B                                         \\
      \textit{Programs}      & c       & ::= & (\textrm{See Fig.~\ref{pi-terms}})
    \end{array}$}}
\begin{figure}[t]
  {\scalebox{\scalef}{$%
        %%\noindent\begin{minipage}{.7\linewidth}
        \begin{array}{rrcll}
          \idc :     & A                     & \iso & A                            & : \idc      \\
          \\
          \identlp : & \zerot + A            & \iso & A                            & : \identrp  \\
          \swapp :   & A + B                 & \iso & B + A                        & : \swapp    \\
          \assoclp : & A + (B + C)           & \iso & (A + B) + C                  & : \assocrp  \\ [1.5ex]
          \identlt : & \onet \times A        & \iso & A                            & : \identrt  \\
          \swapt :   & A \times B            & \iso & B \times A                   & : \swapt    \\
          \assoclt : & A \times (B \times C) & \iso & (A \times B) \times C        & : \assocrt  \\ [1.5ex]
          \absorbr : & ~ \zerot \times A     & \iso & \zerot ~                     & : \factorzl \\
          \dist :    & ~ (A + B) \times C    & \iso & (A \times C) + (B \times C)~ & : \factor
        \end{array}$}}

@@

The denotation of types is straightforward:

\begin{align*}
  \denot{\zerot}     & = \emptyset                  \\
  \denot{\onet}      & = \{ \Acon{tt} \}            \\
  \denot{A + B}      & = \denot{A} \uplus \denot{B} \\
  \denot{A \times B} & = \denot{A} \times \denot{B}
\end{align*}

where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
sets. The denotation of a combinator $c : A \isot B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:

\begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
  \denot{\identlp}           & (\inlv{v})           & = & v                                   \\
  \denot{\identrp}           & v                    & = & \inlv{v}                            \\
  \denot{\swapp}             & (\inlv{v})           & = & \inrv{v}                            \\
  \denot{\swapp}             & (\inrv{v})           & = & \inlv{v}                            \\
  \denot{\assoclp}           & (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
  \denot{\assoclp}           & (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
  \denot{\assoclp}           & (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
  \denot{\assocrp}           & (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
  \denot{\assocrp}           & (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
  \denot{\assocrp}           & (\inrv{v})           & = & \inrv{(\inrv{v})}                   \\
  \denot{\identlt}           & (\Acon{tt} , v)      & = & v                                   \\
  \denot{\identrt}           & v                    & = & (\Acon{tt} , v)                     \\
  \denot{\swapt}             & (v_1 , v_2)          & = & (v_2 , v_1)                         \\
  \denot{\assoclt}           & (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
  \denot{\assocrt}           & ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
  \denot{\dist}              & (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
  \denot{\dist}              & (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
  \denot{\factor}            & (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
  \denot{\factor}            & (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)                  \\
  \denot{\idc}               & v                    & = & v                                   \\
  \denot{(c_1 \fatsemi c_2)} & v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
  \denot{(c_1 \oplus c_2)}   & (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
  \denot{(c_1 \oplus c_2)}   & (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
  \denot{(c_1 \otimes c_2)}  & (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
\end{longtable}

% \begingroup
% \allowdisplaybreaks
% \begin{align*}
% \end{align*}
% \endgroup

@@

\note{change thm:
  1. Each function is a bijection. 2. If there is a 2-combinator, the denotations of the 1-combinators are equal
}
